---
title: 'In Brief - Interactive Maps of Major US Storms: 1996 - 2011'
author: "Mark Blackmore"
date: "August 16, 2017"
output: 
  html_document:
    keep_md: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Introduction
This report explores the NOAA Storm Database and creates interactive maps, in the interest of answering two basic questions:  

1. Across the United States, which types of events have had at least ten million dollars in property damage? 
2. Across the United States, which types of events have had at least ten fatalities? 

Click on a circle or marker to see storm event type.

```{r message=FALSE}
library(tidyverse)

### NOAA Storn Database
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

### Codebook
codebook_url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf"

### Forum Notes on Handling PROPDMGEXP, CROPDMGWXP Variables
forum_url <- "https://www.coursera.org/learn/reproducible-research/discussions/weeks/4/threads/IdtP_JHzEeaePQ71AQUtYw"

exp_vars_url <- "https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html"

#download.file(url, destfile = "StormData.csv")
stormData <- read.csv("StormData.csv", stringsAsFactors = FALSE)
```


```{r}
### Treat downloaded data as read-only
stormData1 <- stormData

## Change event begin date varaiable class from character to date
stormData1$BGN_DATE <- as.Date(stormData$BGN_DATE, format = "%m/%d/%Y")

## Subset to years 1996 - 2011
stormData2 <- stormData1 %>% filter(BGN_DATE >= "1996-01-01")

## Change event type variable from character to factor for further investigation
stormData2$EVTYPE   <- factor(stormData2$EVTYPE)
stormData2$PROPDMGEXP   <- factor(stormData2$PROPDMGEXP)  
stormData2$CROPDMGEXP   <- factor(stormData2$CROPDMGEXP)

## Correct Event Types to NOAA specification
stormData2$EVTYPE[stormData2$EVTYPE=="HURRICANE/TYPHOON"] <- "HURRICANE"
stormData2$EVTYPE[stormData2$EVTYPE=="TYPHOON"] <- "HURRICANE"
stormData2$EVTYPE[stormData2$EVTYPE=="STORM SURGE"] <- "STORM SURGE/TIDE"
stormData2$EVTYPE[stormData2$EVTYPE=="TSTM WIND"] <- "THUNDERSTORM WIND"
stormData2$EVTYPE[stormData2$EVTYPE=="TSTM WIND/HAIL"] <- "THUNDERSTORM WIND"
stormData2$EVTYPE[stormData2$EVTYPE=="MARINE TSTM WIND"] <- "MARINE THUNDERSTORM WIND"
stormData2$EVTYPE[stormData2$EVTYPE=="WILD/FOREST FIRE"] <- "WILDFIRE"
stormData2$EVTYPE[stormData2$EVTYPE=="FOG"] <- "DENSE FOG"
stormData2$EVTYPE[stormData2$EVTYPE=="URBAN/SML STREAM FLD"] <- "FLOOD"
stormData2$EVTYPE[stormData2$EVTYPE=="WINTER WEATHER/MIX"] <- "WINTER WEATHER"
stormData2$EVTYPE[stormData2$EVTYPE=="HEAVY SURF/HIGH SURF"] <- "HIGH SURF"
stormData2$EVTYPE[stormData2$EVTYPE=="TSTM WIND (G45)"] <- "THUNDERSTORM WIND"
stormData2$EVTYPE[stormData2$EVTYPE=="TSTM WIND (G40)"] <- "THUNDERSTORM WIND"
stormData2$EVTYPE[stormData2$EVTYPE=="STRONG WINDS"] <- "HIGH WIND"
stormData2$EVTYPE[stormData2$EVTYPE=="FREEZING RAIN"] <- "SLEET"
stormData2$EVTYPE[stormData2$EVTYPE=="EXTREME WINDCHILL TEMPERATURES"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="COLD/WIND CHILL"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="EXTREME COLD"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="EXTREME WINDCHILL"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="COLD"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="RECORD COLD"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="UNSEASONABLY COLD"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="Cold"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="UNUSUALLY COLD"] <- "EXTREME COLD/WIND CHILL"
stormData2$EVTYPE[stormData2$EVTYPE=="SNOW"] <- "HEAVY SNOW"
stormData2$EVTYPE[stormData2$EVTYPE=="Snow"] <- "HEAVY SNOW"
stormData2$EVTYPE[stormData2$EVTYPE=="EXCESSIVE SNOW"] <- "HEAVY SNOW"
stormData2$EVTYPE[stormData2$EVTYPE=="WIND"] <- "HIGH WIND"
stormData2$EVTYPE[stormData2$EVTYPE=="Heavy Rain"] <- "HEAVY RAIN"
stormData2$EVTYPE[stormData2$EVTYPE=="RECORD RAINFALL"] <- "HEAVY RAIN"
stormData2$EVTYPE[stormData2$EVTYPE=="RECORD WARMTH"] <- "HEAT"
stormData2$EVTYPE[stormData2$EVTYPE=="UNUSUAL WARMTH"] <- "HEAT"
stormData2$EVTYPE[stormData2$EVTYPE=="RECORD HEAT"] <- "HEAT"
stormData2$EVTYPE[stormData2$EVTYPE=="UNSEASONABLY WARM"] <- "HEAT"
stormData2$EVTYPE[stormData2$EVTYPE=="Winter Weather"] <- "WINTER WEATHER"
stormData2$EVTYPE[stormData2$EVTYPE=="FREEZE"] <- "FROST/FREEZE"
stormData2$EVTYPE   <- factor(stormData2$EVTYPE)

## Mutate to common numeric/dollar basis, incorporating exp variables. Rescaled to Millions of Dollars ($000,000)
stormData3 <- stormData2 %>%
  mutate(prop_damage = ifelse(PROPDMGEXP == "", PROPDMG*10^-6, 
    ifelse(PROPDMGEXP == "B", PROPDMG*10^3, 
    ifelse(PROPDMGEXP == "M", PROPDMG*10^0, 
    ifelse(PROPDMGEXP == "K", PROPDMG*10^-3,
    ifelse(PROPDMGEXP ==  0,   PROPDMG*10^-5, PROPDMG)))))) %>%
  mutate(crop_damage = ifelse(CROPDMGEXP == "", CROPDMG*10^-6, 
    ifelse(CROPDMGEXP == "B", CROPDMG*10^3, 
    ifelse(CROPDMGEXP == "M", CROPDMG*10^0, 
    ifelse(CROPDMGEXP == "K", CROPDMG*10^-3,
    ifelse(CROPDMGEXP ==  0,  CROPDMG*10^-5, CROPDMG))))))

## Checking Outliers & Correcting Entries
stormData3$prop_damage[stormData3$prop_damage == 1.15e+05] <- 115

## Correct Entry for Katrina
stormData3$LATITUDE[stormData3$prop_damage == 3.13e+04]  <- 2995
stormData3$LONGITUDE[stormData3$prop_damage == 3.13e+04] <- 9007

```

# Map of Major Storms in 50 States: Over $10M Property Damages

```{r message=FALSE, fig.width=9}
## For 50 States Only use 
stormData4 <- stormData3[(stormData3$STATE %in% state.abb),]

## Storms with over $10M Property Damage
library(leaflet)
stormData5 <- stormData4 %>% filter(LONGITUDE != 0) %>% 
  filter(prop_damage >= 1e+02) %>% mutate(lat = LATITUDE/100, 
    lng = -abs(LONGITUDE)/100) %>% 
  select(lat,lng, EVTYPE)

## Create Interactive Map 
stormDataMap1 <- stormData5  %>% leaflet() %>% addTiles() %>%  
  addMarkers(clusterOptions = markerClusterOptions(), popup = stormData5$EVTYPE)
stormDataMap1
```

# Map of Major Storms in 50 States: Over 10 Fatalities
```{r message=FALSE, fig.width=9}
## Storms with over 10 Fatalities
library(leaflet)
stormData6 <- stormData4 %>% filter(LONGITUDE != 0) %>% 
  filter(FATALITIES >= 10) %>% mutate(lat = LATITUDE/100, 
    lng = -abs(LONGITUDE)/100) %>% 
  select(lat,lng, EVTYPE)

## Create Interactive Map 
stormDataMap2 <- stormData6  %>% leaflet() %>% addTiles() %>%  
  addMarkers(clusterOptions = markerClusterOptions(), popup = stormData5$EVTYPE)
stormDataMap2
```

